Viscous Erosion in Porous Media

### 2018-03-23 docs/methods/erosion_methods.tex
edited section 3.1

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical methods}
\label{s:method}
\todo[inline]{Here on Bryan's sections of editing}
To perform the simulations, we freeze the geometry at each time step,
solve the incompressible Stokes equations, and then update the geometry
according to the erosion model.  The separation between these two
processes is justified since the time scale of erosion is much longer
than that of the flow.

Given this separation of time scales, there are three main numerical
methods that need to be developed.  In Section~\ref{sec:BIE}, we
describe how the BIE~\eqref{eqn:completed_BIE} is solved.  Once the
density function is found, Section~\ref{sec:shearStress} describes how
the shear stress~\eqref{eqn:shearStressLP} is computed.  Then, in
Section~\ref{sec:timeStepping}, a time stepping method for evolving the
interface according to equations~\eqref{eqn:Levo1}
and~\eqref{eqn:thetaevo1} is described. So that we can form stable
simulations with long time horizons, we use numerical methods that
achieve spectral accuracy in space, second-order accuracy in time, and
maintain an equispaced discretization of the bodies.  The boundaries are
represented in Fourier space, giving us a set of discretization points
$\xx_i$ of each body, and all derivatives and integrals (the arclength,
curvature, and mean values) are computed with spectral accuracy using
the Fourier representation.

@@

To numerically evaluate the completed double-layer
potential~\eqref{eqn:completed_BIE}, let $N_\iin$ be the number of
points on each of the $M$ bodies, and $N_\out$ be the number of points
on the outer boundary.  Then, $N = MN_\iin + N_\out$ is the total number
of discretization points of $\bd\Omega$.  The double-layer potential
in~\eqref{eqn:DLP} is discretized with the trapezoid rule as
\begin{align}
  \DD[\eeta](\xx_i) = \sum_{j=1}^{N} K(\xx_i,\xx_j) \eeta(\xx_j) 
      \Delta s_{j}, \quad i=1,\ldots,N,
  \label{eqn:trapDLP}
\end{align}
where $\Delta s_j$ is the arclength term of $\bd\Omega$ at
$\xx_j$, and
\begin{align*}
  K(\xx,\yy) = \frac{1}{\pi} \frac{\rr \cdot \nn}{\rho^2} 
      \frac{\rr \otimes \rr}{\rho^2},
\end{align*}
where $\rr = \xx - \yy$ and $\rho = \|\rr\|$, is the kernel of the
double-layer potential.  The diagonal term of~\eqref{eqn:trapDLP} is
replaced with the limiting value of the double-layer potential
\begin{align*}
  D(\xx,\xx) = -\frac{\kappa(\xx)}{2\pi}(\ss(\xx) \otimes \ss(\xx)),
\end{align*}
where $\kappa(\xx)$ is the curvature and $\ss(\xx)$ is the unit tangent
vector at $\xx \in \bd\Omega$.  The equations for the
Stokeslets~\eqref{eqn:stokeslet} and the rotlets~\eqref{eqn:rotlet} are
also discretized with the trapezoid rule.  Applying this discretization
strategy to~\eqref{eqn:completed_BIE}, the dense linear system that
needs to be solved is
\begin{equation}
  \label{eqn:completed_BIE_discretized}
  \begin{aligned}
    \ff(\xx_i) &= -\frac{1}{2}\ssigma(\xx_i) + \sum_{j=1}^{N} 
      K(\xx_i,\xx_j) \eeta(\xx_j) \Delta s_{j} + 
      \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx_i) \\
      &\hspace{53pt}
      +\sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx_i) 
      +\sum_{\xx_j \in \Gamma} (\nn(\xx_i) \otimes
      \nn(\xx_j))\eeta(\xx_j), \\
    \llambda_\ell &= \sum_{\xx_j \in \gamma_\ell} \eeta(\xx_j) 
      \Delta s_j, \\ 
    \mu_\ell &= \sum_{\xx_j \in \gamma_\ell}
      {(\xx_j - \cc_\ell)}^\perp \cdot \eeta(\xx_j) \Delta s_j,
  \end{aligned}
\end{equation}
for $i=1,\ldots,N$, $\ell=1,\ldots,M$.  Since all the integrands are
smooth and periodic, by using the trapezoid rule, we are guaranteed that
the numerical solutions converges to the true solution with spectral
accuracy.

The linear system~\eqref{eqn:completed_BIE_discretized} is solved
iteratively with the generalized minimal residual method
(GMRES)~\cite{saa-sch1986}.  Since the BIE is second-kind, only a
mesh-independent number of GMRES iterations are
required~\cite{cam-ips-kel-mey-xue1996}.  Therefore, the asymptotic cost
is determined by the cost of performing a matrix-vector multiplication,
and the bulk of the cost is evaluating the $N$-term sum
in~\eqref{eqn:completed_BIE_discretized}.  Naively, evaluating this sum
at all points requires $\bigO(N^2)$ operations.  To reduce the
computational cost, we use the Fast Multipole Method
(FMM)~\cite{gre-rok1987, gre-gre-may1992} to reduce the complexity from
$\bigO(N^2)$ to $\bigO(N)$.  Since the number of required GMRES
iterations is mesh-independent, the overall complexity for
solving~\eqref{eqn:completed_BIE_discretized} is $\bigO(N)$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Shear stress}
\label{sec:shearStress}
We have decomposed the deformation tensor into contributions from three
different terms.  The contributions from the Stokeslets and the rotlets
can be directly evaluated since $\rr \neq 0$ for all $\xx \in \gamma$.
For the shear stress of the double-layer potential, we would like to use
the trapezoid rule. The kernel of the deformation tensor due to
the double-layer potential~\eqref{eqn:shearStressDLP} is 
\begin{align*}
  K_\ssigma(\xx,\yy)\eeta(\yy) &= \frac{1}{2\pi} \left(
    2\frac{\rr \cdot \nn}{\rho^2} \frac{\rr \cdot \eeta}{\rho^2} \II + 
    \frac{\rr \cdot \eeta}{\rho^4} (\nn \otimes \rr + \rr \otimes \nn)
    \right. \\ & \hspace{25pt}+ \left.
    \frac{\rr \cdot \nn}{\rho^4} (\eeta \otimes \rr + \rr \otimes \eeta) - 
    8\frac{(\rr \cdot \nn)(\rr \cdot \eeta)}{\rho^6}(\rr \otimes \rr)
  \right).
\end{align*}
Unfortunately, this kernel has a singularity that is asymptotic to
$\bigO(\rho^{-2})$ when $\yy \rightarrow \xx$, and this is too strong
for the trapezoid rule.  The singularity strength can be reduced since
the velocity corresponding to a constant density function is
constant~\cite{poz1992}.  Therefore, the deformation tensor of the
double-layer potential is equivalent to
\begin{align*}
  \ssigma^{\DD}(\xx) = \int_{\bd\Omega}K_\ssigma(\xx,\yy)
      (\eeta(\yy) - \eeta(\xx)) \, ds_{\yy}, \quad \xx \in \bd\Omega.
\end{align*}
This expression for the deformation tensor has a singularity that is
asymptotic to $\bigO(\rho^{-1})$ when $\yy \rightarrow \xx$.  Finally,
to achieve spectral accuracy, we apply odd-even
integration~\cite{sid-isr1988}.  Therefore, the deformation tensor of
the double-layer potential at a discretization point $\xx_i$ on body
$\gamma_\ell$, with $i$ being odd, is approximated as
\begin{align}
  \ssigma^{\DD}(\xx_i) = \sum_{\xx_j \notin \gamma_\ell}
    K_\ssigma(\xx_i,\xx_j) \eeta(\xx_j) \Delta s_j + 
  \sum_{\substack{\xx_j \in \gamma_\ell \\ j \: \mathtt{even}}}
    K_\ssigma(\xx_i,\xx_j) (\eeta(\xx_j) - \eeta(\xx_i)) \Delta s_j,
  \label{eqn:stressOddEven}
\end{align}
and a similar approximation is used when $i$ is even.  The deformation
tensor due to the Stokeslets and rotlets are given in
equation~\eqref{eqn:shearStressSR}.  Finally, the jump in the
deformation tensor~\eqref{eqn:shearStressJump} is computed with Fourier
differentiation.  Finally, the shear stress is computed by multiplying the
deformation tensor by the normal vector, and taking the inner product
with the tangent vector.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical methods}
\label{s:method}
To perform numerical simulations, we exploit the difference
between the erosion and fluid time scales.  Since erosion occurs
much slower, we freeze the geometry at each time
step, solve the incompressible Stokes equations, and then update the
geometry according to the erosion model.  Therefore, there are three
main numerical methods that need to be developed.  In
Section~\ref{sec:BIE}, we describe how the BIE~\eqref{eqn:completed_BIE}
is solved.  Once the density function is found,
Section~\ref{sec:shearStress} describes how the shear
stress~\eqref{eqn:shearStressLP} is computed.  Then, in
Section~\ref{sec:timeStepping}, a time stepping method for evolving the
interface according to equations~\eqref{eqn:Levo1}
and~\eqref{eqn:thetaevo1} is described.

To form stable simulations with long time horizons, we use numerical
methods that achieve spectral accuracy in space, second-order accuracy
in time, and maintain an equispaced discretization of the bodies.  The
boundaries are represented in Fourier space, giving us a set of
discretization points $\xx_i$ of each body, and all derivatives and
integrals (the arclength, curvature, and mean values) are computed with
spectral accuracy using the Fourier representation.

@@

To numerically compute the completed double-layer
potential~\eqref{eqn:completed_BIE}, we discretize each of the $M$
bodies with  $N_\iin$ points, and the outer boundary with $N_\out$
points.  Then, $N = MN_\iin + N_\out$ is the total number of
discretization points of $\bd\Omega$.  The double-layer potential
in~\eqref{eqn:DLP} is discretized with the trapezoid rule as
\begin{align}
  \DD[\eeta](\xx_i) = \sum_{j=1}^{N} K(\xx_i,\xx_j) \eeta(\xx_j) 
      \Delta s_{j}, \quad i=1,\ldots,N,
  \label{eqn:trapDLP}
\end{align}
where $\Delta s_j$ is the arclength term of $\bd\Omega$ at
$\xx_j$, and
\begin{align*}
  K(\xx,\yy) = \frac{1}{\pi} \frac{\rr \cdot \nn}{\rho^2} 
      \frac{\rr \otimes \rr}{\rho^2},
\end{align*}
where $\rr = \xx - \yy$ and $\rho = \|\rr\|$, is the kernel of the
double-layer potential.  The diagonal term of~\eqref{eqn:trapDLP} is
replaced with the limiting value of the double-layer potential
\begin{align*}
  D(\xx,\xx) = -\frac{\kappa(\xx)}{2\pi}(\ss(\xx) \otimes \ss(\xx)),
\end{align*}
where $\kappa(\xx)$ is the curvature and $\ss(\xx)$ is the unit tangent
vector at $\xx \in \bd\Omega$.  The equations for the
Stokeslets~\eqref{eqn:stokeslet} and the rotlets~\eqref{eqn:rotlet} are
also discretized with the trapezoid rule.  Applying this discretization
strategy to~\eqref{eqn:completed_BIE}, the dense linear system that
needs to be solved is
\begin{equation}
  \label{eqn:completed_BIE_discretized}
  \begin{aligned}
    \ff(\xx_i) &= -\frac{1}{2}\ssigma(\xx_i) + \sum_{j=1}^{N} 
      K(\xx_i,\xx_j) \eeta(\xx_j) \Delta s_{j} + 
      \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx_i) \\
      &\hspace{53pt}
      +\sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx_i) 
      +\sum_{\xx_j \in \Gamma} (\nn(\xx_i) \otimes
      \nn(\xx_j))\eeta(\xx_j), \\
    \llambda_\ell &= \sum_{\xx_j \in \gamma_\ell} \eeta(\xx_j) 
      \Delta s_j, \\ 
    \mu_\ell &= \sum_{\xx_j \in \gamma_\ell}
      {(\xx_j - \cc_\ell)}^\perp \cdot \eeta(\xx_j) \Delta s_j,
  \end{aligned}
\end{equation}
for $i=1,\ldots,N$, $\ell=1,\ldots,M$.  We have applied the trapezoid
rule since all the integrands are smooth and periodic, and this
guarantees that the numerical solutions converges to the true solution
with spectral accuracy.

The linear system~\eqref{eqn:completed_BIE_discretized} is solved
iteratively with the generalized minimal residual method
(GMRES)~\cite{saa-sch1986}.  Since the BIE is second-kind, only a
mesh-independent number of GMRES iterations are
required~\cite{cam-ips-kel-mey-xue1996}.  Therefore, the asymptotic cost
is determined by the cost of performing a matrix-vector multiplication,
and the bulk of the cost is evaluating the $N$-term sum
in~\eqref{eqn:completed_BIE_discretized}. We use the Fast Multipole
Method (FMM)~\cite{gre-rok1987, gre-gre-may1992} so that the cost of
solving~\eqref{eqn:completed_BIE_discretized} is $\bigO(N)$ operations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Shear stress}
\label{sec:shearStress}
\todo[inline]{Bryan is here on editing}
We have decomposed the deformation tensor into contributions from three
different terms.  The contributions from the Stokeslets and the rotlets
can be directly evaluated since $\rr \neq 0$ for all $\xx \in \gamma$.
For the shear stress of the double-layer potential, we would like to use
the trapezoid rule. The kernel of the deformation tensor due to
the double-layer potential~\eqref{eqn:shearStressDLP} is 
\begin{align*}
  K_\ssigma(\xx,\yy)\eeta(\yy) &= \frac{1}{2\pi} \left(
    2\frac{\rr \cdot \nn}{\rho^2} \frac{\rr \cdot \eeta}{\rho^2} \II + 
    \frac{\rr \cdot \eeta}{\rho^4} (\nn \otimes \rr + \rr \otimes \nn)
    \right. \\ & \hspace{25pt}+ \left.
    \frac{\rr \cdot \nn}{\rho^4} (\eeta \otimes \rr + \rr \otimes \eeta) - 
    8\frac{(\rr \cdot \nn)(\rr \cdot \eeta)}{\rho^6}(\rr \otimes \rr)
  \right).
\end{align*}
Unfortunately, this kernel has a singularity that is asymptotic to
$\bigO(\rho^{-2})$ when $\yy \rightarrow \xx$, and this is too strong
for the trapezoid rule.  The singularity strength can be reduced since
the velocity corresponding to a constant density function is
constant~\cite{poz1992}.  Therefore, the deformation tensor of the
double-layer potential is equivalent to
\begin{align*}
  \ssigma^{\DD}(\xx) = \int_{\bd\Omega}K_\ssigma(\xx,\yy)
      (\eeta(\yy) - \eeta(\xx)) \, ds_{\yy}, \quad \xx \in \bd\Omega.
\end{align*}
This expression for the deformation tensor has a singularity that is
asymptotic to $\bigO(\rho^{-1})$ when $\yy \rightarrow \xx$.  Finally,
to achieve spectral accuracy, we apply odd-even
integration~\cite{sid-isr1988}.  Therefore, the deformation tensor of
the double-layer potential at a discretization point $\xx_i$ on body
$\gamma_\ell$, with $i$ being odd, is approximated as
\begin{align}
  \ssigma^{\DD}(\xx_i) = \sum_{\xx_j \notin \gamma_\ell}
    K_\ssigma(\xx_i,\xx_j) \eeta(\xx_j) \Delta s_j + 
  \sum_{\substack{\xx_j \in \gamma_\ell \\ j \: \mathtt{even}}}
    K_\ssigma(\xx_i,\xx_j) (\eeta(\xx_j) - \eeta(\xx_i)) \Delta s_j,
  \label{eqn:stressOddEven}
\end{align}
and a similar approximation is used when $i$ is even.  The deformation
tensor due to the Stokeslets and rotlets are given in
equation~\eqref{eqn:shearStressSR}.  Finally, the jump in the
deformation tensor~\eqref{eqn:shearStressJump} is computed with Fourier
differentiation.  Finally, the shear stress is computed by multiplying the
deformation tensor by the normal vector, and taking the inner product
with the tangent vector.
